---
title: 'Chicken Metformin: Preliminary Report'
author: "Emily Van Syoc"
date: "12/15/2021"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center",
                      fig.width = 9,
                      fig.height = 7)
knitr::opts_knit$set(root.dir = "/Users/epb5360/OneDrive - The Pennsylvania State University/Shared-Trello-Projects/Chicken Metformin/chicken-metf/")

## read chunks for Wu
knitr::read_chunk("/Users/epb5360/OneDrive - The Pennsylvania State University/Shared-Trello-Projects/Chicken Metformin/chicken-metf/preliminary-report-1-2022.R")

```

## Preliminary Data Report for chicken metformin project

## Data overview

```{r getData}

```

After filtering and removing contaminants, there are 267 bacterial genera in 14 unique phyla.

```{r relAbun}

```

## Statistical models

After consulting with Justin Silverman, I made 2 comparisons:  
1. Compare control and Treatment 4 -- *is there a difference caused by metformin*  
2. Compare times -- *is there a difference between times*  
3. Compare metformin treatments at all times -- *is there a dose response*  

## Alpha Diversity

Alpha diversity is called "richness" and generally compares the number of taxa between groups. Due to the compositional nature of microbiome data, we use different indices to compare alpha diversity.

I calculated Observed taxa (only for exploratory analysis, not statistical comparisons), Shannon's H index, and Simpson's diversity.

In Shannon's diversity, there is only a difference between times 2 and 3. There is no difference in Simpson's diversity.

```{r AlphaDiversity}

```

## Beta Diversity

Beta diversity is the microbial community composition or structure. It refers to the presence/absence of different taxa as well as their relative abundances between groups. 

Beta diversity is compared by transforming the OTU table into a distance matrix using a specific statistic (in this case, we use Bray-Curtis distance). Then, we conduct a permutational ANOVA on the distance matrix to compare groups (called an "adonis" test). These ANOVAs are visualized in an ordination plot.

Beta diversity is commonly more responsive than alpha diversity. We see strong differences in all comparisons. 

```{r BetaDiversity}

```

## Differential relative abundance

Differential relative abundance is calculated with a modified version of log2 fold change with a package called DESeq. These comparisons were made at the Genus level.


There are significantly different (Benjamani-Hochberg corrected) genera in each comparison. However, the "Control over Time" comparison shows that there is also natural variation in the microbiome over time. 

Notably, some taxa differ by over a 20 log2 fold change, which is very strong.

```{r RelativeAbundance}


```

